PLANETESIMAL ACCRETION AT SHORT ORBITAL PERIODS

### 2021-11-24 main.tex
More content

In figure \ref{fig:timescales}, we show ratio between the relaxation and collision timescale for a population of equal-mass planetesimals as a function of orbital period. Here, the encounter velocity is described by $v = \left( 5/8 \left< e^2 \right>^{1/2} + \left< i^2 \right>^{1/2} \right)^{1/2} v_{k}$, where $\left< e^2 \right>^{1/2}$ and $\left< i^2 \right>^{1/2}$ are the eccentricity and inclination dispersions of the planetesimal disk and $v_{k}$ is the Keplerian velocity. For simplicity, we assume that $\left< e^2 \right>^{1/2} = 2\left< i^2 \right>{1/2}$ \citep{ida93a} and that the eccentricity dispersion is constant with orbital period. The eccentricity dispersion is described in units scaled by the Hill factor $m_{pl}/ 3 M_{star}$ such that $e_{h} = 1$ corresponds to the boundary between shear and dispersion dominated encounters. The horizontal dashed line indicates where $t_{relax} = t_{coll}$. The timescale criterion for oligarchic growth is only satisfied in regions where the disk is sufficiently dynamically cold and the orbital period is sufficiently long. In section \ref{sec:results} we will explore the behavior and outcome of planetesimal accretion in regions where this criterion is \textit{not} satisfied.

@@

@@

We begin by exploring a parameter study of planetesimal accretion, motivated by two questions raised in section \ref{sec:theory}. 1. Does runaway growth still operate when the condition that $v \ll v_{esc}$ is not satisfied? 2. How does planetesimal accretion proceed when $\alpha$ is no longer small?

To answer these questions, we run a series of simulations in which a narrow annulus of planetesimals orbits a star. The values of $\alpha$ and $\beta = v / v_{esc}$ are varied individually. 4000 planetesimals with individual masses of $5 \times 10^{23}$ g are placed with semimajor axes randomly drawn between 0.95 and 1.05 AU about a 1 $M_{\odot}$ star. The argument of perihelion $\omega$, longitude of ascending node $\Omega$, and mean anomaly M for each body is drawn from a uniform distribution $\in [0, 2 \pi)$. The inclination dispersion $\left< i^{2} \right>^{1/2}$ is chosen to be half the eccentricity dispersion $\left< e^{2} \right>^{1/2}$ \citep{ida93a}.

In the `nominal' case, we give the bodies a bulk density of 3 g cm$^{-3}$, and $\left< e^{2} \right>^{1/2} = 4 e_{h}$, where $e_{h} = e a/r_{h}$ (define this in the previous section instead) which corresponds to $\alpha = 3.6 \times 10^{-2}$ and $\beta = 3.4 \times 10^{-3}$. These parameters are chosen to match the initial conditions of \citet{kokubo98}, which gave rise to oligarchic growth. To vary the value of $\alpha$, we alter the bulk density of the particles. In the high-$\alpha$ case, the bulk density is reduced by a factor of $\sim$ 7100, which produces $\alpha = 1$. To vary $\beta$, the eccentricity dispersion is increased. For the high-$\beta$ case, $\left< e^{2} \right>^{1/2}$ is increased to $1500 e_{h}$, which corresponds to $\beta = 15,000$.
% Should I choose beta = 1 instead? I did beta >> 1 to see if inelastic mergers will cool things off. Havent decided if that's important here

In all cases, the simulations are evolved with a base timestep of 1.7 days, which corresponds to 3\% of an orbital dynamical time $\sqrt{a^3/G M_{*}}$. Due to the vastly differing growth timescales in each case, a simulation is stopped when the growth of the most massive body flattens out. In figure \ref{fig:alpha_beta}, we show the a-e distribution of bodies in the initial (blue) and final (orange) snapshots from each of the 4 simulations. The size of the points indicates the relative masses of the bodies. Only in the case of small $\alpha$ does a residual population of dynamically hot planetesimals develop. The lack of high eccentricity planetesimals in the large $\alpha$ case suggests that most encounters instead result in accretion. In the case of large $\beta$, the protoplanets and remaining planetesimals end up in a dynamically cool state, compared to the initial conditions. Here, the only mechanism that can remove energy are the merger events which result in an inelastic collision.

@@

To investigate whether any of these planetesimal disks underwent runaway growth, we examine the time evolution of the maximum and mean masses in each simulation. The ratio $m_{max}/\left< m \right>$ is plotted in figure \ref{fig:alpha_beta_evo}. On this plot, a positive slope indicates that the growth rate is accelerating with mass, which is evident in all four cases. Even with large $\beta$, runaway growth still appears to operate for some duration of time.

@@

@@

>>>>

In figure \ref{fig:timescales}, we show ratio between the relaxation and collision timescale for a population of equal-mass planetesimals as a function of orbital period. Here, the encounter velocity is described by $v = \left( 5/8 \left< e^2 \right>^{1/2} + \left< i^2 \right>^{1/2} \right)^{1/2} v_{k}$, where $\left< e^2 \right>^{1/2}$ and $\left< i^2 \right>^{1/2}$ are the eccentricity and inclination dispersions of the planetesimal disk and $v_{k}$ is the Keplerian velocity. For simplicity, we assume that $\left< e^2 \right>^{1/2} = 2\left< i^2 \right>{1/2}$ \citep{ida93a} and that the eccentricity dispersion is constant with orbital period. The eccentricity dispersion is described in units scaled by the Hill factor $\left( m_{pl}/ 3 M_{star} \right)^{1/3}$ such that $e_{h} = 1$ corresponds to the boundary between shear and dispersion dominated encounters. The horizontal dashed line indicates where $t_{relax} = t_{coll}$. The timescale criterion for oligarchic growth is only satisfied in regions where the disk is sufficiently dynamically cold and the orbital period is sufficiently long. In section \ref{sec:results} we will explore the behavior and outcome of planetesimal accretion in regions where this criterion is \textit{not} satisfied.

@@

We also introduce a second dimensionless quantity, which relates the physical size of the bodies to the velocity state of the system

\begin{equation}\label{eq:beta}
	\beta = \frac{r_{pl}}{r_{g}}.
\end{equation}

\noindent where $r_{g} = G m_{pl} / v^{2}$ is the gravitational radius of a body. Encounters between bodies inside of a distance of $r_{g}$ result in significant deflections of their trajectories. It should be noted that the gravitational focusing enhancement factor $v^{2}/v_{esc}^{2}$ is equal to 1 for $\beta = 1$. In the case where $r_{g}$ is smaller than the size of a planetesimal, the gravitational focusing enhancement factor will be between 0 and 1.

@@

We begin by presenting a parameter study of planetesimal accretion, motivated by two questions raised in section \ref{sec:theory}. 1. Does runaway growth still operate when the condition that $v \ll v_{esc}$ is not satisfied? 2. How does planetesimal accretion proceed when the planetesimals themselves occupy a significant fraction of their Hill spheres?

To answer these questions, we run a series of simulations in which a narrow annulus of planetesimals orbits a star. The values of $\alpha$ and $\beta$ are varied individually. 4000 planetesimals with individual masses of $5 \times 10^{23}$ g are placed with semimajor axes randomly drawn between 0.95 and 1.05 AU about a 1 $M_{\odot}$ star. The argument of perihelion $\omega$, longitude of ascending node $\Omega$, and mean anomaly M for each body is drawn from a uniform distribution $\in [0, 2 \pi)$. The inclination dispersion $\left< i^{2} \right>^{1/2}$ is chosen to be half the eccentricity dispersion $\left< e^{2} \right>^{1/2}$ \citep{ida93a}.

In the `nominal' case, we give the bodies a bulk density of 3 g cm$^{-3}$, and $\left< e^{2} \right>^{1/2} = 4 e_{h}$, where $e_{h} = e a/r_{h}$ (define this in the previous section instead) which corresponds to $\alpha = 3.6 \times 10^{-2}$ and $\beta = 3.4 \times 10^{-3}$. These parameters are chosen to match the initial conditions of \citet{kokubo98}, which gave rise to oligarchic growth. To vary the value of $\alpha$, we alter the bulk density of the particles. In the high-$\alpha$ case, the bulk density is reduced by a factor of $\sim$ 7100, which produces $\alpha = 1$. To vary $\beta$, the eccentricity dispersion is increased. For the high-$\beta$ case, $\left< e^{2} \right>^{1/2}$ is increased to $1500 e_{h}$, which corresponds to $\beta = 15,000$.
% Should I choose beta = 1 instead? I did beta >> 1 to see if inelastic mergers will cool things off. Havent decided if that's important here
% Another concern: using too large of a beta makes the apo-peri distances larger than the width of the annulus

In all cases, the simulations are evolved with a base timestep of 1.7 days, which corresponds to 3\% of an orbital dynamical time $\sqrt{a^3/G M_{*}}$. Due to the vastly differing growth timescales in each case, a simulation is stopped when the growth of the most massive body flattens out. In figure \ref{fig:alpha_beta}, we show the a-e distribution of bodies in the initial (blue) and final (orange) snapshots from each of the 4 simulations. The size of the points indicates the relative masses of the bodies. Only in the case of small $\alpha$ does a residual population of dynamically hot planetesimals develop. The lack of high eccentricity planetesimals in the large $\alpha$ case suggests that most encounters instead result in accretion. In the case of large $\beta$, the protoplanets and remaining planetesimals end up in a dynamically cool state, compared to the initial conditions. (Are the inelastic collisions playing a significant role here?)

In figure \ref{fig:alpha_beta_mass}, we show the mass distribution of bodies from the final snapshot in each of the four cases. In addition to leaving fewer residual planetesimals, the large $\alpha$ simulations produce significantly larger embryos. Despite the vastly different encounter velocities of each population of bodies, the initial size of $\beta$ appears to have no significant effect on the final distribution of masses.

@@

To investigate whether any of these planetesimal disks underwent runaway growth, we examine the time evolution of the maximum and mean masses in each case. The ratio $m_{max}/\left< m \right>$ is plotted in figure \ref{fig:alpha_beta_evo}. On this plot, a positive slope indicates that the growth rate is accelerating with mass, which is evident in all four cases. Even with a large value of $\beta$, which means that the effective collision cross section is very near the geometric value, runaway growth still appears to operate. This seems to imply that, as bodies collide and grow, the relative increase in their gravitational focusing factors, rather than its absolute size, is what drives the system towards runaway growth. Although larger encounter velocities will lengthen the growth timescales, this mode of growth appears to be inevitable, so long as gravity is the dominant force in the system.

Additionally, these results suggest that the value of $\alpha$, which is a function of only the initial conditions, controls the qualitative outcome of accretion. Across most of a planet-forming disk, $\alpha$ is small, and frequent gravitational encounters between the growing bodies will allow oligarchic growth to operate. At shorter orbital periods, however, the Hill sphere of a body is no longer mostly empty space, and it appears as if collisions play a more significant role. In this regime, we observe that runaway growth still commences, but nearly all of the planetesimals are swept up by the forming protoplanets, rather than being scattered onto higher eccentricity orbits, where they would otherwise remain as a remnant of the early stages of planet formation \citep{kokubo98, kokubo00}.

\begin{table}
\begin{center}
\caption{Summary of Full Disk Simulations Run}
\begin{tabular}{lllll} \hline \hline
Name     & A & $\alpha$ &  &  \\ \hline
fdHi                 & 100                                                         & 1.5                              &  &  \\
fdHiShallow     & 100                                                        & 0.5                              &  &  \\
fdHiSteep       & 100                                                         & 2.5                              &  &  \\
fdLo                & 1                                                             & 1.5                              &  &  \\ \hline
\end{tabular}
\label{tab:sims}
\end{center}
\end{table}

@@

Motivated by the apparent dependence of accretion modes on $\alpha$, we next investigate whether this highly efficient, non-oligarchic growth mode should be expected to operate in nature. Given the dearth of short period terrestrial planets observed around M stars (citation), we model a series of much wider disks, which span from 1 to 100 days in orbital period orbiting a late-type M star of mass 0.08 $M_{\odot}$. For a population of planetesimals with a bulk density of 3 g cm$^{-3}$, this orbital period range corresponds to $\alpha \in (0.7, 0.05)$. By simultaneously modeling a broad range of orbital periods, we can determine the critical value of $\alpha$ that divides these two modes of accretion, and also explore how the oligarchic/non-oligarchic accretion boundary affects the resulting distribution of protoplanets.

Four wide disk simulations are run in total (see table \ref{tab:sims}). In each case, the solid surface density follows a power law profile

\begin{equation}
	\Sigma(r) = \left( \frac{M_{*}}{M_{\odot}} \right) A \Sigma_{mmsn} \left( \frac{r}{1 AU} \right)^{-\alpha},
\end{equation}

\noindent where $M_{*}$ is the mass of the central star, $\Sigma_{MMSN} = 10$ g cm$^{-2}$ is surface density of the minimum-mass solar nebula \citep{hayashi81} at 1 AU, and $A$ is an enhancement factor. In the first case (fdHi), we model a disk that follows the surface density

More stuff about initial conditions, timestepping issues, gas drag

@@

\begin{figure}
\begin{center}
    \includegraphics[width=0.5\textwidth]{figures/alpha_beta_mass.png}
    \caption{Caption goes here.\label{fig:alpha_beta_mass}}
\end{center}
\end{figure}
